---
title: "Heatmaps"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape2)
```

Required packages

```{r eval = FALSE}
library(ggplot2)
library(reshape2)
```




## 1) What ia a heatmap?

A heat map (or heatmap) is a literal way of visualizing a table of numbers, where 
you substitute the numbers with colored cells. In other words, a heatmap is a 
graphical representation of data where values are depicted by color. 

Why heatmaps are popular?

- They make it easy to visualize tabular data and understand it at a glance.
It's useful for finding highs and lows and sometimes, patterns.

- The idea is to take advantage of the human color vision to detect/observe
patterns that otherwise are very difficult to uncover just by looking at the
numeric values in a table.

- Because heatmaps encode numbers with colors, the visual display can be very 
colorful, which if plotted following good design principles, is pleasent to 


### 1.1) Why are they called heatmaps?

As mentioned above, when using a heatmap, numerical data gets graphically 
represented in a way that the value of each data point is indicated using colors. 
The most commonly used color scheme used in heatmap visualization is the 
__warm-to-cool color scheme__, with the warm colors representing 
high-value data points and the cool colors representing low-value data points.
Hence the reason for the name _heatmap_.




## 2) Intuition of heatmaps

```{r Xmatrix, echo = FALSE}
num_rows = 4
num_cols = 3
num_total = num_rows * num_cols

X = matrix(1:num_total, nrow = num_rows, ncol = num_cols)
rownames(X) = paste0("row", 1:num_rows)
colnames(X) = paste0("col", 1:num_cols)
```

Let us start by describing the graphing steps behind a heatmap. To do this, 
consider the following matrix `X` 

```{r echo = FALSE, comment = ""}
X
```

As you can tell, `X` is of size $`r num_rows` \times `r num_cols`$, that is, 
it has `r num_rows` rows and `r num_cols` columns, for a total of `r num_total`
entries (or cells). To create this matrix in R, use the following code:

```{r Xmatrix}
```

For visualization purposes, we could take the numeric values in the entries
of `X`, and map them to different colors. For example, we could map a value 
of 1 to color "blue", 2 to "turquoise", 3 to "green"`, etc. And then, graph
the data in `X` as an image in which each entry is plotted with its respective
assigned color.

As an example, we use the `rainbow()` function that creates a ramp pallete
with as many colors as the indicated input number.

```{r}
some_colors <- rainbow(num_total)
X_colors <- matrix(some_colors, nrow = num_rows, ncol = num_cols)
dimnames(X_colors) <- dimnames(X)
X_colors
```

The basic idea of a heatmap is to plot a rectangular grid, with each cell colored 
according to the mapped value, something like the following diagram

```{r echo = FALSE, fig.align='center', , out.width='70%'}
mycols <- c(
  X_colors[4,],
  X_colors[3,],
  X_colors[2,],
  X_colors[1,])

xcols = rep(1:num_cols, times = num_rows)
yrows = rep(1:num_rows, each = num_cols)
plot(xcols, yrows, axes = FALSE, type = "n",
     xlim = c(0.5,3.5), ylim = c(1,4.5))
axis(side = 2, at = 1:num_rows, labels = num_rows:1, las = 1)
axis(side = 3, at = 1:num_cols, labels = 1:num_cols)
symbols(xcols, yrows, squares = rep(0.1, num_total), 
        inches = FALSE, add = TRUE, bg = mycols)
```

In fact, a proper heatmap will look like the following graph:

```{r echo = FALSE, fig.align='center', out.width='50%'}
Y = rbind(X[4,], X[3,], X[2,], X[1,])
heatmap(Y, Rowv = NA, Colv = NA, scale = "none", col = X_colors,
        labRow = rev(rownames(X)))
```


R provides the function `heatmap()`, that creates default heatmaps. Now,
keep in mind that the default order in which rows and columns are organized by
`heatmap()` differs from the above example. Turning off arguments `Rowv`, `Colv`,
and `scale`, we get the following image:

```{r fig.align='center', out.width='50%'}
heatmap(X, Rowv = NA, Colv = NA, scale = "none", col = X_colors)
```

It is the same data, the difference in the heatmaps has to do with the order 
of the rows.

Because data values are transformed to color scale, a heatmap is said to be a 
"false colored image".



## 3) Plotting heatmaps with R

There are a multiple numbers of R packages and functions for drawing interactive 
and static heatmaps, including:

- `heatmap()` from `"base"` packages: Draws a simple heatmap

- `heatmap.2()` from "gplots"` package: Draws an enhanced heatmap compared to 
the R base function.

- `pheatmap()` from `"pheatmap": Draws pretty heatmaps and provides more control 
to change the appearance of heatmaps.

- `Heatmap()` from `"ComplexHeatmap"` R/Bioconductor package: Draws, annotates 
and arranges complex heatmaps (very useful for genomic data analysis)

- and many more.

We will only cover the `heatmap()` and `heatmap.2()` functions in this tutorial.


### 3.1) Example in R

Let's use the data frame `mtcars`. First we order values by `mpg` (miles per gallon)

```{r}
dat <- mtcars[order(mtcars$mpg), ]
head(dat, n = 5)
```

The data is in data frame format, but it has to be a data matrix to make a 
heatmap

```{r}
dat_matrix <- data.matrix(dat)
class(dat_matrix)
```

Once the data is in matrix form, we can use the function `heatmap()` as
follows. The main input is the data (matrix). In the code below we use the 
arguments `Rowv = NA` and `Colv = NA` to display a heatmap without clustering
the rows and columns. In addition, we scecify that the data should be scaled 
taking into account the columns (i.e. the variables). This scaling or normalizing
operation transforms each column in a way that they have mean zero and unit 
standard deviation (i.e. standardized data)

```{r out.width='80%', fig.align='center'}
nba_heatmap <- heatmap(
  dat_matrix, 
  Rowv = NA, 
  Colv = NA, 
  #col = cm.colors(256), 
  scale = "column", 
  margins = c(5,10))
```


#### Changing Colors

You can change the color scheme with the argument `col`. For example, we can
choose a _cyam-magenta_ palette with `cm.colors()`

```{r out.width='80%', fig.align='center'}
nba_heatmap <- heatmap(
  dat_matrix, 
  Rowv = NA, 
  Colv = NA, 
  col = cm.colors(256), 
  scale = "column", 
  margins = c(5,10))
```

Maybe you want a different color scheme. For example, you could use 
more heat-looking colors with `heat.colors()`:

```{r out.width='80%', fig.align='center'}
nba_heatmap <- heatmap(
  dat_matrix, 
  Rowv=NA, 
  Colv=NA, 
  col = heat.colors(256), 
  scale = "column", 
  margins=c(5,10))
```


### 3.2) Basic heatmap with `"ggplot2"`

What about graphing heatmaps with `"ggplot2"`? Although this package does not
have a dedicated `heatmap()`-like function, it is possible to graph basic 
heatmaps with the geometric object function `geom_tile()`.

In this part of the tutorial you need to load the following packages

```{r eval = FALSE}
library(ggplot2)    # for graphics
library(reshape2)   # for data reshaping
```

THe graphing process of heatmaps with `"ggplot2"` is not straightforward as
with base `heatmap()`. One thing to remember is that `ggplot()` expects to 
receive a data frame as its main input. The good news is that `mtcars` is 
already in data frame format. The bad news is that we need to do some data 
reshaping (wrangling). This is where `:reshape2"` comes very handy.

First we need to `melt()` the data frame. For convenience purposes, we will 
create a "copy" of `mtcars`, and then play with our copy instead of directly
modifying `mtcars`:

```{r}
dat_cars <- mtcars
dat_cars$car <- rownames(mtcars)
dat_melt <- melt(dat_cars)
```

Once its melted, we can use `geom_tile()`

```{r out.width='80%', fig.align='center'}
ggplot(data = dat_melt, aes(x = variable, y = car, fill = value)) +
  geom_tile()
```


#### Scaling Variables

The variables with the largest variance will dominate the scale. To avoid 
this phenomenon, it is typical to transform variables into comparable units

```{r}
mtcars_scaled <- as.data.frame(scale(mtcars))
mtcars_scaled$car <- rownames(mtcars_scaled)
dat_melt <- melt(mtcars_scaled)
head(dat_melt)
```

Let's replot the heatmap

```{r out.width='80%', fig.align='center'}
ggplot(data = dat_melt, aes(x = variable, y = car, fill = value)) +
  geom_tile()
```

Let's now try a different color palette

```{r out.width='80%', fig.align='center'}
ggplot(data = dat_melt, aes(x = variable, y = car, fill = value)) +
  geom_tile() +
  scale_fill_distiller(palette = 'PiYG')
```

Here's another option

```{r out.width='80%', fig.align='center'}
ggplot(data = dat_melt, aes(x = variable, y = car, fill = value)) +
  geom_tile() +
  scale_fill_distiller(palette = 'Spectral')
```

We can try another scaling option such as the _minmax_ which converts any
scale into a 0-1 range:

```{r}
minmax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

mt_min <- apply(mtcars, 2, min)
mt_max <- apply(mtcars, 2, max)
mt_range <- mt_max - mt_min

dat_min_centered <- sweep(mtcars, MARGIN = 2, STATS = mt_min, FUN = "-")
dat_minmax <- sweep(dat_min_centered, MARGIN = 2, STATS = mt_range, FUN = "/")

# confirming minmax scale
apply(dat_minmax, 2, min)
apply(dat_minmax, 2, max)
```

```{r}
dat_minmax$car <- rownames(mtcars)
dat_melt <- melt(dat_minmax)
head(dat_melt)
```

```{r out.width='80%', fig.align='center'}
ggplot(data = dat_melt, aes(x = variable, y = car, fill = value)) +
  geom_tile() +
  scale_fill_distiller(palette = 'PiYG')
```




## Clustered Heatmaps

Instead of having the horizontal axis represent levels or values of a single 
variable, it is a common variation to have it represent measurements of 
different variables or metrics. If we set the vertical axis as individual 
observations, we end up with something resembling a standard data table, where 
each row is an observation and the columns the entity’s value on each measured 
variable.

This type of heatmap is sometimes known as a clustered or clustering heatmap, 
since the goal of this kind of chart is to build associations between both the 
data points and their features. We want to see which individuals are similar or 
different from each other, with a similar objective for variables. Analysis 
tools that construct this type of heatmap will usually implement clustering as 
part of their process. This use case is found in areas like the biological 
sciences, such as when studying similarities in gene expression across individuals.

Heat maps allow us to simultaneously visualize clusters of samples and features. 
First hierarchical clustering is done of both the rows and the columns of the 
data matrix. The columns/rows of the data matrix are re-ordered according to 
the hierarchical clustering result, putting similar observations close to each 
other. The blocks of "high" and "low" values are adjacent in the data matrix. 
Finally, a color scheme is applied for the visualization and the data matrix is 
displayed. Visualizing the data matrix in this way can help to find the 
variables that appear to be characteristic for each sample cluster.




## Examples with `mtcars`

Basic example

```{r out.width='80%', fig.align='center'}
dat <- scale(mtcars)

# scaling data by rows
heatmap(dat, scale = "row")
```

The arguments are

- `dat` is a numeric matrix

- `scale` a character indicating if the values should be centered and scaled in 
either the row direction or the column direction, or none. Allowed values are 
in `c("row", "column", "none")`. Default is `"row"`.

In the plot above, high values are in red and low values are in yellow.


```{r out.width='80%', fig.align='center'}
# scaling data by colums
heatmap(dat, scale = "column")
```


It's possible to specify a color palette using the argument col, which can be 
defined as follow:

Using custom colors:

```{r}
# red, white, blue
rwb <- colorRampPalette(c("red", "white", "blue"))(256)

heatmap(dat, scale = "row", col = rwb)
```


### Enhanced heatmaps with `heatmap.2()`

The function `heatmap.2()` [in `"gplots"` package] provides many extensions to 
the standard R `heatmap()` function presented in the previous section.

```{r}
# install.packages("gplots")
library("gplots")
heatmap.2(dat, scale = "none", col = rwb, 
          trace = "none", density.info = "none")
```

