---
title: "PCA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(FactoMineR)
library(ggplot2)
```

```{r eval = FALSE}
# required packages
library(FactoMineR)
library(ggplot2)
```


# PCA on Genomic Data

This example uses the data `poulet` ("Chicken" in French) that comes in the R 
package `"FactoMineR"` (by F. Husson, S. Le, J. Pages), and was kindly donated 
by Prof. Sandrine Lagarrigue (Agrocampus, Rennes, France).

```{r}
# load data
data(poulet)

dim(poulet)
```


## Context

The data relates to 43 chickens having undergone one of the six following 
diet treatments:

- `N`: normal diet
- `J16`: fasting for 16 hours
- `J16R5`: fasting for 16 hours then refed for 5 hours
- `J16R16`: fasting for 16 hours then refed for 16 hours
- `J48`: fasting for 48 hours
- `J48R24`: fasting for 48 hours then refed for 24 hours

```{r}
summary(poulet$Diet)
```

At the end of the diet, the genes were analyzed using DNA chips, and the 
expression of 7407 genes retained for all the chickens. A biologist selected
the most pertinent genes, since at the beginning, more than 20,000 genes were
identified by DNA chips. The data were then preprocessed in a standard manner
for DNA chips (normalizing, eliminating the chip effect, etc.).

The data table `poulet` is a data frame with 43 rows (chickens) and 7407
columns. The first column `Diet` refers to the diet treatment, and the rest of
the columns refer to the genes.

The aim of the study is to see whether the genes are expressed differently 
depending on the situation of stress (i.e. the type of diet) which the chicken
is subject to. More precisely, it may be interesting to see how long the 
chicken needs to be refed after fasting before it returns to a normal state, 
i.e. a state comparable to the state of a chicken with a normal diet. Might some
genes be unexpressed during fasting and overexpressed during feeding?


## Performing PCA

R has several functions (and packages) that allows us to performa a principal
components analysis.

One good starting option is the `prcomp()` function.

Should the variables be standardized? In the `poulet` data, the variables are
already standardized to provide the same influence to each gene.

```{r}
pca <- prcomp(poulet[ ,-1], scale. = FALSE)
```

`prcomp()` returns an object of class `"prcomp"`, which is an R list containing
the following elements:

```{r}
names(pca)
```

- `sdev`: stands for _standard deviation_ and it contains the standard 
deviations of the principal components

- `rotation`: is the matrix of so-called _loadings_, that is the coefficients
that let you form the principal components from the input variables; 
it is called rotation because this is a _rotation matrix_ in linear algebra sense.

- `center`: is a vector with the mean of each input variable

- `scale`: is a logica value indicating whether the input variables are standardized
or not

- `x`: is a matrix containing whose columns are the principal components, 
_aka_ PCA scores.


### What to look for?

- Inspect the values in `sdev`
- Graph scatterplots of PCs in `x`
- Graph scatterplots of loadings in `rotation`


### PC's Variation

Recall that `sdev` contains the standard deviations of the PCs. If we square
these values, then we obtain the variances. The variances tells us how much
variability (i.e. how much information) a given PC is _capturing_ from the 
total variation in the data.

```{r}
# standard devitions of PCs
head(pca$sdev)
```

```{r}
# variances of PCs
head(pca$sdev^2)
```

For convenience purposes, we can assemble a table with three columns: the
variance of the PCs, the variance in percentage (proportion), and the cumulative proportions:

```{r}
pc_var <- data.frame(
  variance = round(pca$sdev^2, 4),
  percent = round(100 * pca$sdev^2 / sum(pca$sdev^2), 4),
  cum_percent = 100 * cumsum(pca$sdev^2 / sum(pca$sdev^2))
)

head(pc_var, n = 5)
tail(pc_var, n = 5)
```

Notice that the first PC has a variance of about 213.37, which accounts for the
28.12% of total variability in the data. In turn, the second PC represents 
10.90% of the variation in the data, and so on. Combined, the first two PCs
capture about 40% of the variability in the data. Also, the last row in `pc_var` 
has a variance of 0.

This means that we obtained 42 dimensions at most, which corresponds to the 
number of individuals minus 1 (and not the total number of variables): the 
42 chickens are therefore in a space with 42 dimensions at most.

Another way to think about what is happening is like this. The data contains 43 
chickens on which we observe 4706 variables (genes). Geometrically, the data 
points lie---theoretically---in a 43 dimensional space. If the genetic information
in each chicken was unique, then each chicken would provide 
$1/43 \times 100\% \approx 2.3255\%$ of the variation in the data.
However, as you can tell from the above output,
the first PC is actually capturing 28.12% of the variation! This tells us that
there is a substantial amount of redundant (duplicated) genetic information in 
the chickens. 

We can see this better with a barplot of the variance proportions. We add a 
dashed red horizontal line at 100/43, to indicate the theoretical proportion of
unique variability that each chicken would account for, provided no redundant
information was shared among them.

```{r}
barplot(pc_var$percent, las = 1, ylim = c(0, 30), border = NA)
lines(1:length(pc_var$percent), pc_var$percent)
abline(h = 100/43, col = "red", lty = 2)
```


### PC Scatterplots

Let's now inspect a scatterplot of PCs

```{r}
plot(pca$x[ ,1], pca$x[ ,2], las = 1)
abline(h = 0, v = 0, col = "gray80")
```

```{r echo = FALSE, eval = FALSE}
plot(pca$x[ ,1], pca$x[ ,2], las = 1,
     col = poulet$Diet)
text(pca$x[ ,1], pca$x[ ,2], las = 1,
     labels = poulet$Diet, col = factor(poulet$Diet))
abline(h = 0, v = 0, col = "gray80")
```

We can color code the chickens based on type of diet

```{r}
pc_dat <- cbind.data.frame(
  diet = poulet$Diet,
  pca$x
)
head(pc_dat)
```

```{r}
ggplot(data = pc_dat, aes(x = PC1, y = PC2, color = diet)) +
  geom_vline(xintercept = 0, color = "gray90") +
  geom_hline(yintercept = 0, color = "gray90") +
  geom_point() +
  geom_text(label = rownames(pc_dat), size = 2) +
  theme_classic()
```

The principal plane separates the chickens into two subgroups:

- those that have undergone intense stress (48 hours of fasting), are greatly
scattered on the right side of the plot

- those that been subject to less intense stress (16 hours of fasting), are 
more concentrated on the left side, and located closer to the origin

Furthermore, the second PC separates the chickens into three groups: chickens
that have undergone intense stress but that have not been refed afterwards
(`J48R24`), chickens that have undergone intense stress and that haven beed
refed afterwards (`J48R24`), and the other chickens.

Chickens which have been refed tend to recover from intense stress and their
health tends to be similar to that of normal chickens.

However, 24 hours of feeding does not seem to be enough for the state of the
chicken to completely return to normal. This indicates that some genes are
specific to a state of intense stress, and some genes are overexpressed under
stress while others are underexpressed.



## Scatterplot of loadings

So far we've inspected the amount of variation captured by each PC, and also
the main scatterplot between PC1 and PC2. But we should also take a look at 
waht's happening with the _loadings_.

Think of a loading as the weight or "importance" that a variable has behind
a given PC. For instance, PC1 is calculated with the 4706 loading values 
$w_{1,1}, w_{2,1}, \dots, w_{4706,1}$ in the first column of `rotation`:

$$
\texttt{PC1} = w_{1,1} \texttt{A4GALT} + w_{2,1} \texttt{A4GNT} + \dots +  w_{7405,1} \texttt{ZZEF111} + w_{7406,1} \texttt{ZZZ3}
$$

In turn, PC2 is calculated with the 4706 loading values 
$w_{1,2}, w_{2,2}, \dots, w_{4706,2}$ in the second column of `rotation`:

$$
\texttt{PC2} = w_{1,2} \texttt{A4GALT} + w_{2,2} \texttt{A4GNT} + \dots +  w_{7405,2} \texttt{ZZEF111} + w_{7406,2} \texttt{ZZZ3}
$$

One way to explore loadings is by graphing a scatterplot of the columns in the
`rotation` matrix. For example, here's the plot of the loadings for the first
and the second PCs:

```{r}
load_dat <- as.data.frame(pca$rotation)

ggplot(data = load_dat, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.3)
```

Each point represents the loadings of PC1 and PC2 for a given variable.
As you can tell, the cloud of points is somewhat unclear with a large proportion
of points vering overlapped (due to the large number of variables).

Another way to explore the structure of the loadings is by looking at the 
correlations between the input variables and the PCs

Let $\mathbf{X}$ be the $43 \times 7406$ matrix of varibales, and $\mathbf{PC}$ 
the $43 \times 43$ matrix of PCs. The correlation matrix between variables and
PCs is given by:

$$
Cor(\mathbf{X}, \mathbf{PC}) = \begin{bmatrix}
cor(X_1, \text{PC}_1) & cor(X_1, \text{PC}_2) & \cdots & cor(X_1, \text{PC}_{43}) \\
cor(X_2, \text{PC}_1) & cor(X_2, \text{PC}_2) & \cdots & cor(X_2, \text{PC}_{43}) \\
\vdots & \vdots & \ddots & \vdots \\
cor(X_{7406}, \text{PC}_1) & cor(X_{7406}, \text{PC}_2) & \cdots & cor(X_{7406}, \text{PC}_{43}) \\
\end{bmatrix}
$$

In R you can calculate this matrix as follows. We convert the output into a 
data frame for graphing purposes via `ggplot()`:

```{r}
cor_dat <- as.data.frame(cor(poulet[ ,-1], pca$x))
```

Instead of directly plotting the loading values, we can graph the correlation
values:

```{r}
ggplot(data = cor_dat, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.3)
```

This cloud of points is regularly distributed and does not need to be commented
(but nonetheless had to be verified).

We can look for those genes (variables) that are the most correlated with each 
PC. Somewhat arbitrarily, we may begin looking for the top-10 largest correlations
and see what variables (genes) are responsible for this.

For example, with respect to PC1 the top 10 largest correlations come from 
the following genes:

```{r}
tail(cor_dat[order(abs(cor_dat[,1])),1,drop = FALSE], n = 10)
```

For the PC2 we have:

```{r}
tail(cor_dat[order(abs(cor_dat[,2])),2,drop = FALSE], n = 10)
```

After identifying those genes that are highly correlated with PC1 and PC2
(and possibly more PCs), we could ask a biologist/geneticist to analyze why
these particular genes are under or overexpressed.


